---
title: "Astros' Sign Stealing Effectiveness"
author: "Luis, Edgar, Adam"
date: "2025-04-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lubridate)
library(knitr)
library(kableExtra)
library(ggthemes)
library(ggrepel)
library(scales)
library(plotly)
library(broom)
library(patchwork)
```

# Introduction

## Background


The 2017 Houston Astros' won the World Series, but their victory was later tainted when they were accused of cheating their way to victory. The team would relay pitch signals via trash can bangs. A bang typically indicated an off-speed pitch, while no bang indicated a fastball.

In 2020, the MLB confirmed these allegations following an investigation. The scandal resulted in the suspension of manager A.J. Hinch and general manager.

## Purpose of the Study

We aim to quantify the impact of their sign-stealing system and if it actually imporved batting performance. We can combine data on trash bang occurrences and available player statistics.


##Questions

1. Did batters perform better during at-bats with trash can bangs?
2. Did a higher frequency of bangs per game translate to more wins?
3. How many home runs were hit when bangs were detected versus without bangs?
4. Which players received the most benefit from the banging system?
5. Did the effectiveness of bangs change over the course of the season?
6. Was there a difference in impact based on pitch type?

## Data Sources

- **Bang Data**: A dataset compiled by a fan who reviewed footage of all Astros' home games from the 2017 season. This data is available at [signstealingscandal.com](https://signstealingscandal.com/).
- **Player Statistics**: Game-by-game and pitch-by-pitch data from [baseball-reference.com](https://www.baseball-reference.com/) and [retrosheet.org](https://www.retrosheet.org/)
- **Advanced Metrics**: Additional analysis and visualizations from [crawfishboxes.com](https://www.crawfishboxes.com/2020/1/29/21114099/advanced-data-on-the-astros-cheating-how-often-main-culprits-what-effect)

# Data Preparation

```{r data-import}
# Import trash can bang data
bangs_data <- read_csv("data/astros_bangs.csv")

# Import player performance 
player_stats <- read_csv("data/astros_batting_data.csv")

# Import game results
game_results <- read_csv("data/astros_game_results.csv")

#Import player game logs
player_files <- list.files(path = "data/game_logs", pattern = "*.csv", full.names = TRUE)
player_stats <- data.frame()

for (file in player_files) {
  player_data <- read_csv(file)

  player_name <- gsub(".csv", "", basename(file))
  player_data$Player <- player_name
  
 
  player_stats <- bind_rows(player_stats, player_data)
}



head(bangs_data)
head(player_stats)
head(game_results)
head(player_files)
head(player_stats)
```



```{r data-cleaning}
# First, inspect the date formats to ensure proper conversion
head(bangs_data$game_date)
head(player_stats$Date)
head(game_results$Date)

# Clean and transform the bangs data
bangs_clean <- bangs_data %>%
  # Using as.Date() which is more flexible with date formats
  mutate(game_date = as.Date(game_date)) %>%
  mutate(
    month = month(game_date, label = TRUE),
    home_away = if_else(grepl("vs", opponent), "Home", "Away")
  )

# Clean player stats data
player_stats_clean <- player_stats %>%
  mutate(game_date = as.Date(Date))

# Summarize bangs by game and batter
bangs_summary <- bangs_data %>%
  group_by(game_date, batter) %>%  
  summarize(
    total_bangs = n(),
    .groups = 'drop'
  )

# Join player stats with bangs data
# Make sure we have the correct columns selected for joining
player_stats_with_bangs <- player_stats_clean %>%
  left_join(
    bangs_clean %>% 
      select(game_date, batter, pitch_category), # Select columns needed from bangs data
    by = c("game_date" = "game_date", "Player" = "batter")
  )

# Create player-level summary of bangs
player_bangs_summary <- bangs_clean %>%
  group_by(batter) %>%
  summarize(
    total_bangs = n(),
    avg_bangs_per_game = n() / n_distinct(game_date),
    .groups = 'drop'
  ) %>%
  arrange(desc(total_bangs))

# Convert the Date column in game_results with explicit format handling
game_results_clean <- game_results %>%
  
  # For MM/DD/YYYY format:
  mutate(game_date = as.Date(Date, format = "%m/%d/%Y")) %>%
 
  select(-Date)  

# Create game-level summary and join with game results
game_bangs_summary <- bangs_clean %>%
  group_by(game_date, opponent) %>%
  summarize(
    total_bangs = n(),
    .groups = 'drop'
  ) %>%
  # Join using game_date now that it exists in both dataframes
  left_join(game_results_clean, by = "game_date") %>%
  # Use R and RA columns for determining wins/losses
  mutate(result = if_else(R > RA, "Win", "Loss"))

# Check for NA values in the cleaned datasets
cat("NA values in bangs_clean:", sum(is.na(bangs_clean)), "\n")
cat("NA values in player_stats_clean:", sum(is.na(player_stats_clean)), "\n")
cat("NA values in game_bangs_summary:", sum(is.na(game_bangs_summary)), "\n")

# View the structure of the cleaned datasets
glimpse(bangs_clean)
glimpse(player_stats_clean)
glimpse(game_bangs_summary)
```


## Overall Bang Frequency

Let's first look at how frequently the trash can banging system was used throughout the 2017 season.

```{r monthly-bang-frequency}
monthly_bangs <- bangs_clean %>%
  group_by(month) %>%
  summarize(
    total_bangs = n(),
    games = n_distinct(game_date),
    bangs_per_game = total_bangs / games
  )

ggplot(monthly_bangs, aes(x = month, y = bangs_per_game)) +
  geom_col(fill = "#002D62") +
  geom_text(aes(label = round(bangs_per_game, 1)), vjust = -0.5) +
  labs(
    title = "Average Trash Can Bangs per Game by Month (2017)",
    x = "Month",
    y = "Average Bangs per Game"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

```{r game-by-game-bangs}
# Plot bangs by game
ggplot(game_bangs_summary, aes(x = game_date, y = total_bangs, fill = result)) +
  geom_col() +
  scale_fill_manual(values = c("Win" = "#002D62", "Loss" = "#EB6E1F")) +
  labs(
    title = "Trash Can Bangs by Game (2017 Season)",
    x = "Game Date",
    y = "Number of Bangs",
    fill = "Game Result"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 90, hjust = 1)
  )
```

## Player-Specific Bang Analysis

Which players received the most help from the trash can banging system?

```{r player-bang-frequency}
# Plot bangs by player (top 10)
top_players <- player_bangs_summary %>%
  top_n(10, total_bangs)

ggplot(top_players, aes(x = reorder(batter, total_bangs), y = total_bangs)) +
  geom_col(fill = "#002D62") +
  geom_text(aes(label = total_bangs), hjust = -0.2) +
  labs(
    title = "Players Receiving the Most Trash Can Bangs (2017)",
    x = "Player",
    y = "Total Bangs"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

```{r bangs-by-lineup-position}
# Data Line up by position
bangs_by_position <- bangs_clean %>%
  group_by(lineup_position) %>%
  summarize(
    total_bangs = n(),
    players = n_distinct(batter)
  ) %>%
  mutate(bangs_per_player = total_bangs / players)

ggplot(bangs_by_position, aes(x = as.factor(lineup_position), y = bangs_per_player)) +
  geom_col(fill = "#002D62") +
  labs(
    title = "Average Bangs per Player by Lineup Position",
    x = "Lineup Position",
    y = "Bangs per Player"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

## Game Situation Analysis

```{r bangs-by-inning}
# Analyze bangs by inning
bangs_by_inning <- bangs_clean %>%
  group_by(inning) %>%
  summarize(total_bangs = n())

ggplot(bangs_by_inning, aes(x = as.factor(inning), y = total_bangs)) +
  geom_col(fill = "#002D62") +
  labs(
    title = "Trash Can Bangs by Inning",
    x = "Inning",
    y = "Number of Bangs"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

```{r bangs-by-score-differential}
# Analyze bangs by score differential
bangs_by_score <- bangs_clean %>%
  mutate(score_diff = astros_score - opponent_score) %>%
  group_by(score_diff) %>%
  summarize(total_bangs = n())

ggplot(bangs_by_score, aes(x = score_diff, y = total_bangs)) +
  geom_col(fill = "#002D62") +
  labs(
    title = "Trash Can Bangs by Score Differential",
    x = "Astros Lead/Deficit",
    y = "Number of Bangs"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

```{r home-away-comparison}
# Compare home vs away performance
# Note: The sign stealing system was primarily used at home
home_away_stats <- player_stats_clean %>%
  group_by(home_away) %>%
  summarize(
    games = n_distinct(game_date),
    batting_avg = sum(hits) / sum(at_bats),
    obp = sum(hits + walks + hbp) / sum(at_bats + walks + hbp + sacrifice_flies),
    slg = sum(total_bases) / sum(at_bats),
    ops = obp + slg
  )

# Tidy format for plotting
home_away_tidy <- home_away_stats %>%
  pivot_longer(
    cols = c(batting_avg, obp, slg, ops),
    names_to = "metric",
    values_to = "value"
  )

ggplot(home_away_tidy, aes(x = metric, y = value, fill = home_away)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Home" = "#002D62", "Away" = "#EB6E1F")) +
  labs(
    title = "Astros Batting Performance: Home vs. Away (2017)",
    x = "Metric",
    y = "Value",
    fill = "Location"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

# Performance Impact Analysis

## Team Performance

Let's analyze whether the trash can bangs had a measurable impact on team performance.

```{r wins-and-bangs}
# Analyze relationship between bangs and wins
bangs_vs_wins <- game_bangs_summary %>%
  group_by(result) %>%
  summarize(
    games = n(),
    total_bangs = sum(total_bangs),
    avg_bangs_per_game = total_bangs / games
  )

ggplot(bangs_vs_wins, aes(x = result, y = avg_bangs_per_game, fill = result)) +
  geom_col() +
  scale_fill_manual(values = c("Win" = "#002D62", "Loss" = "#EB6E1F")) +
  geom_text(aes(label = round(avg_bangs_per_game, 1)), vjust = -0.5) +
  labs(
    title = "Average Trash Can Bangs in Wins vs. Losses",
    x = "Game Result",
    y = "Average Bangs per Game"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

```{r correlation-bangs-runs}
# Analyze correlation between bangs and runs scored
ggplot(game_bangs_summary, aes(x = total_bangs, y = astros_score)) +
  geom_point(aes(color = result), size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", se = TRUE) +
  scale_color_manual(values = c("Win" = "#002D62", "Loss" = "#EB6E1F")) +
  labs(
    title = "Relationship Between Trash Can Bangs and Runs Scored",
    x = "Number of Bangs in Game",
    y = "Astros Runs Scored",
    color = "Game Result"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )

# Calculate correlation coefficient
correlation <- cor.test(game_bangs_summary$total_bangs, game_bangs_summary$astros_score)
print(correlation)
```

## Individual Performance

```{r bang-vs-no-bang-performance}
# Compare batting performance with and without bangs
performance_comparison <- player_stats_clean %>%
  group_by(has_bang) %>%
  summarize(
    at_bats = sum(at_bats),
    hits = sum(hits),
    home_runs = sum(home_runs),
    batting_avg = hits / at_bats,
    hr_rate = home_runs / at_bats,
    obp = sum(hits + walks + hbp) / sum(at_bats + walks + hbp + sacrifice_flies),
    slg = sum(total_bases) / sum(at_bats),
    ops = obp + slg
  )

# Create a tidy format for plotting
performance_tidy <- performance_comparison %>%
  pivot_longer(
    cols = c(batting_avg, hr_rate, obp, slg, ops),
    names_to = "metric",
    values_to = "value"
  )

ggplot(performance_tidy, aes(x = metric, y = value, fill = has_bang)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "#002D62", "FALSE" = "#EB6E1F"), 
                    labels = c("TRUE" = "With Bang", "FALSE" = "Without Bang")) +
  labs(
    title = "Astros Batting Performance With vs. Without Bangs",
    x = "Metric",
    y = "Value",
    fill = "Bang Status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

```{r player-performance-difference}
# Calculate performance difference by player (with vs without bangs)
player_difference <- player_stats_clean %>%
  group_by(batter, has_bang) %>%
  summarize(
    at_bats = sum(at_bats),
    hits = sum(hits),
    batting_avg = hits / at_bats
  ) %>%
  pivot_wider(
    names_from = has_bang,
    values_from = c(at_bats, hits, batting_avg),
    names_glue = "{.value}_{has_bang}"
  ) %>%
  mutate(
    avg_difference = batting_avg_TRUE - batting_avg_FALSE,
    sufficient_sample = (at_bats_TRUE >= 20 & at_bats_FALSE >= 20)
  ) %>%
  filter(sufficient_sample) %>%
  arrange(desc(avg_difference))

# Plot top 10 players by improvement
top_improvers <- player_difference %>%
  top_n(10, avg_difference)

ggplot(top_improvers, aes(x = reorder(batter, avg_difference), y = avg_difference)) +
  geom_col(fill = "#002D62") +
  geom_text(aes(label = round(avg_difference, 3)), hjust = -0.2) +
  labs(
    title = "Players with Greatest Batting Average Improvement with Bangs",
    subtitle = "Minimum 20 at-bats in each condition",
    x = "Player",
    y = "Batting Average Difference (With Bang - Without Bang)"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12)
  )
```

## Pitch Type Analysis

```{r bang-effectiveness-by-pitch}
# Analyze effectiveness of bangs by pitch type
pitch_effectiveness <- pitch_data %>%
  filter(!is.na(has_bang)) %>%
  group_by(pitch_type, has_bang) %>%
  summarize(
    pitches = n(),
    swings = sum(swing),
    hits = sum(hit),
    swing_rate = swings / pitches,
    hit_rate = hits / swings
  )

# Plot hit rate by pitch type and bang status
ggplot(pitch_effectiveness, aes(x = pitch_type, y = hit_rate, fill = has_bang)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("TRUE" = "#002D62", "FALSE" = "#EB6E1F"), 
                    labels = c("TRUE" = "With Bang", "FALSE" = "Without Bang")) +
  labs(
    title = "Hit Rate by Pitch Type and Bang Status",
    x = "Pitch Type",
    y = "Hit Rate (Hits/Swings)",
    fill = "Bang Status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )
```

# Statistical Analysis

Let's perform statistical tests to determine if the differences we observed are statistically significant.

```{r t-test-batting-avg}
# Perform t-test comparing batting average with and without bangs
t_test_result <- t.test(
  player_stats_clean$batting_avg[player_stats_clean$has_bang == TRUE],
  player_stats_clean$batting_avg[player_stats_clean$has_bang == FALSE]
)

print(t_test_result)

# Table to display the results
t_test_table <- tibble(
  Metric = "Batting Average",
  With_Bang = mean(player_stats_clean$batting_avg[player_stats_clean$has_bang == TRUE], na.rm = TRUE),
  Without_Bang = mean(player_stats_clean$batting_avg[player_stats_clean$has_bang == FALSE], na.rm = TRUE),
  Difference = With_Bang - Without_Bang,
  P_Value = t_test_result$p.value
)

kable(t_test_table, 
      caption = "T-Test Results: Batting Average With vs. Without Bangs",
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

```{r regression-analysis}
# Perform regression analysis
# Model to predict batting average based on bang status and other factors
regression_model <- lm(batting_avg ~ has_bang + opponent + inning + outs + runners_on, 
                       data = player_stats_clean)

model_summary <- summary(regression_model)
print(model_summary)

# Table of model coefficients
model_coefs <- tidy(regression_model) %>%
  filter(term %in% c("(Intercept)", "has_bangTRUE")) %>%
  select(term, estimate, std.error, p.value)

kable(model_coefs, 
      caption = "Regression Model Coefficients",
      col.names = c("Term", "Estimate", "Std. Error", "P-Value"),
      digits = 3) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

# Advanced Analysis Based on Crawfish Boxes Data

```{r crawfish-boxes-recreation}
# Recreate some of the analyses from crawfishboxes.com
# Heatmap of bang frequency by count (balls and strikes)
bang_heatmap <- bangs_clean %>%
  group_by(balls, strikes) %>%
  summarize(
    total_bangs = n(),
    at_bats = n_distinct(at_bat_id)
  ) %>%
  mutate(bang_rate = total_bangs / at_bats)

ggplot(bang_heatmap, aes(x = as.factor(strikes), y = as.factor(balls), fill = bang_rate)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "#002D62") +
  labs(
    title = "Trash Can Bang Rate by Count",
    x = "Strikes",
    y = "Balls",
    fill = "Bang Rate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(size = 12)
  )

# Analyze effectiveness against different pitchers
pitcher_effectiveness <- bangs_clean %>%
  group_by(opposing_pitcher) %>%
  summarize(
    total_bangs = n(),
    at_bats = n_distinct(at_bat_id),
    bang_rate = total_bangs / at_bats
  ) %>%
  filter(at_bats >= 10) %>%
  arrange(desc(bang_rate))

# Plot top 10 pitchers targeted
top_targeted_pitchers <- pitcher_effectiveness %>%
  top_n(10, bang_rate)

ggplot(top_targeted_pitchers, aes(x = reorder(opposing_pitcher, bang_rate), y = bang_rate)) +
  geom_col(fill = "#002D62") +
  geom_text(aes(label = sprintf("%.2f", bang_rate)), hjust = -0.2) +
  labs(
    title = "Pitchers Most Frequently Targeted by Trash Can Bangs",
    subtitle = "Minimum 10 at-bats",
    x = "Pitcher",
    y = "Bang Rate (Bangs per At-Bat)"
  ) +
  coord_flip() +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12)
  )
```

# Discussion of Findings

- Did the data show a statistically significant improvement in batting performance when bangs were used?
- Which players benefited the most from the system?
- Were there particular game situations where the bangs were more effective?
- How do your findings align with or contradict previous analyses, such as those from crawfishboxes.com?


# Conclusion

## Summary of Impact

Our analysis sought to quantify whether the trash can banging system used by the Houston Astros in 2017 provided a measurable advantage in batting performance. Based on our findings, we can draw several conclusions:


